---

title: "class notes"
date: "8/30/22"
author: "Magnus Wood"
output: html_notebook
---

Basics of R Markdown are found [here]. (https://rmarkdown.rstudio.com/lesson-8.html)

Data types in R:
* character
* numeric
* integer
* logical
* complex
* raw

Only data of a single type may be stored in a vector. If two data types are present in the same vector, R will change the data type for that vector to a less-restrictive type that can handle both. Character seems common as the least-restrictive. 


str() - object related information
typeof() - what is the objects type(low-level)?
  length() - how long is it? What about two-dimensional objects?
  attributes() - does it have any metadata?

## Data Structures in R
* atomic vector
* list
* array/matrix
* data frame
* factors

### vectors

Vector is most common data type in R and can be either atomic or lists

- Matrix: extension of numeric or character vetors. Not a separate type, just a vector with multiple dimensions
- List: act as containers. Not rest to single mode, can mix data types. 
- Data Frame: tabular data structure. rectangular list.
- Factors: similar to the nomical/ordinal categorical data types. Factors can be ordered or not to reflect this differences. A factor has set levels. 



```{r numericVector}
num_vec <- c(123.1, 123, 1583, 989, 343)
str(num_vec)
typeof(num_vec)
length(num_vec)
attributes(num_vec)

cat("\nsum of numeric vector: ", sum(num_vec), "\n")

res_str <- sprintf("sum of numeric vector: % s", sum(num_vec))
print(res_str)
```
```{r logicalVector}
print("logical vectors can be added together too.")
log_vec <- c(T, T,FALSE, T, TRUE)
str(log_vec)
typeof(log_vec)
length(log_vec)
attributes(log_vec)

cat("\nsum of vector: ", sum(log_vec), "\n")

res_str <- sprintf("sum of  vector: % s", sum(log_vec))
print(res_str)
```

```{r}

```

