---
title: 'AFS505 HW2: Data in R, Fall 2022'
author: "Magnus Wood"
date: "9-6-22"
output:
  html_document:
    df_print: paged
---

## Question 1
###(a) Crop vector
Create the vector shown below:
[1] “apples” “wheat” “soybeans” “potatoes” “grapes” “cherries” “almonds” “hay”
[9] “hops”
 
```{r}
ag_vec <- c("apples", "wheat", "soybeans", "potatoes", "grapes", "cherries", "almonds", "hay", "hops")

ag_vec
```

### (b) Address crop vector
Replace soybeans with oranges.
Use indexing to create a subset of the crop vector with only major crops grown in Washington.
Use a logical vector to create a subset of the crop vector with only major crops grown in Washington.
```{r}
ag_vec[3] <- "oranges"

ag_vec

wa_log <- c(T, T, F, T, T, T, F, T, T)

ag_vec[wa_log]

```

### (c) Reorder WA crop vector
Save the Washington major crop vector (subset of original vector).
Reorder the Washington major crop vector as shown: 
[1] “apples” “grapes” “cherries” “potatoes” “hops” “wheat” “hay”
Use the order() function to sort the Washington major crop vector in alphabetical order.

```{r}
# save subset as new vector
wa_ag_vec <- ag_vec[wa_log]
cat("subset:", wa_ag_vec, "\n\n")

# sort according to given order
ag_order <- factor(wa_ag_vec, 
                   levels=c("apples", "grapes", "cherries", 
                            "potatoes", "hops", "wheat"), 
                   ordered=T)

cat("ordered: ", wa_ag_vec[order(ag_order)], "\n\n")

# alphabetical order
cat("alpha order: ", wa_ag_vec[order(wa_ag_vec)])
      
```

## Question 2
### (a) address matrix with positive indices
Create the matrix below. 
      [,1] [,2] [,3] [,4] [,5]
 [1,]    1    2    3    4    5
 [2,]    6    7    8    9   10
 [3,]   11   12   13   14   15
 [4,]   16   17   18   19   20

```{r}
# create the matrix
x_mat <- matrix(c(1, 6, 11, 16, 2, 7, 12, 17, 3, 8, 13, 18, 4, 9, 14, 19, 5, 10, 15, 20), nrow = 4, ncol=5)
x_mat
```
Extract the following subset: 
      [,1] [,2]
 [1,]    8    9
 [2,]   13   14
 [3,]   18   19

```{r}
mat_sub1 <- x_mat[2:4, 3:4]
mat_sub1
```

Extract the following subset:
 [1]  6  8 10

```{r}
mat_sub2 <- x_mat[2, c(1, 3, 5)]
mat_sub2
```
 
### (b) subset matrix with negative indices
Extract the following subset: 
      [,1] [,2] [,3] [,4] [,5]
 [1,]    1    2    3    4    5
 [2,]    6    7    8    9   10
 [3,]   16   17   18   19   20

```{r}

mat_sub3 <- x_mat[-3,]
mat_sub3
```

Extract the following subset: 
      [,1] [,2] [,3]
 [1,]    1    2    4
 [2,]   11   12   14
 [3,]   16   17   19

```{r}
##      [,1] [,2] [,3] [,4] [,5]
## [1,]    1    2    3    4    5
## [2,]    6    7    8    9   10
## [3,]   11   12   13   14   15
## [4,]   16   17   18   19   20
mat_sub4 <- x_mat[-2, c(-3, -5)]
mat_sub4
```

### (c) replace values in matrix
Create a duplicate matrix for each step of these exercises so the original matrix remains unaltered.
      [,1] [,2] [,3] [,4] [,5]
 [1,]    1    2    3    4    5
 [2,]    6    7    8    9   10
 [3,]   11   12   13   14   15
 [4,]   16   17   18   19   20

Replace 9, 14 and 19 with 88.
```{r}
x_mat1 <- x_mat
x_mat1[c(1, 3, 4), 4] <- 88
x_mat1

```


Replace 9, 14 and 19 with 23, 24, 25.

```{r}
x_mat2 <- x_mat
x_mat2[c(1, 3, 4), 4] <- c(23, 24, 25)
x_mat2

```

Remove the value of 8.
```{r}
x_mat3 <- x_mat
x_mat3[2, 3] <- NA
x_mat3
```

