---
title: 'Homework 3 '
author: "Magnus Wood"
output:
  html_document:
    df_print: paged
---


## Question 1 - simple lists
### (a) List1

Create list1 shown below:
```
## [[1]]
## [1] "Soils 502"
## 
## [[2]]
## [1] 1 3 4 5
## 
## [[3]]
##      [,1] [,2]
## [1,]    1    2
## [2,]    3    4
## [3,]    5    6
```
```{r}
list1 <- list("Soils 502", c(1,3,4,5), matrix(seq(1:6), nrow=3, ncol=2, byrow=TRUE))

list1
```

Extract the following from list1: 
```
##      [,1] [,2]
## [1,]    1    2
## [2,]    3    4
## [3,]    5    6
```

```{r}
list1[[3]]
```

Extract the following from list1: 
```
## [[1]]
## [1] "Soils 502"
## 
## [[2]]
##      [,1] [,2]
## [1,]    1    2
## [2,]    3    4
## [3,]    5    6
## 
## [[1]]
## [1] "Soils 502"
## 
## [[2]]
##      [,1] [,2]
## [1,]    1    2
## [2,]    3    4
## [3,]    5    6
```
```{r}
list1.1 <- list1[c(1, 3)]
list1.1

list1.2 <- list1[c(1, 3)]
list1.2
```
Why twice? Typo?

###(b) list2
Create the following list:
Why aren’t grand slam wins structured as a data frame or matrix?
```
## $name
## $name$Last
## [1] "Williams"
## 
## $name$First
## [1] "Serena"
## 
## 
## $age
## [1] 36
## 
## $record
##   wins losses 
##    801    136 
## 
## $grand.slam.wins
## $grand.slam.wins$Australian.Open
## [1] 2003 2005 2007 2009 2010 2015 2017
## 
## $grand.slam.wins$French.Open
## [1] 2002 2013 2015
## 
## $grand.slam.wins$Wimbledon
## [1] 2002 2003 2009 2010 2012 2015 2016
## 
## $grand.slam.wins$US.Open
## [1] 1999 2002 2008 2012 2013 2014
```

```{r}
# list of mixed data structures, including sublists and named vectors
list2 <- list(name = list(Last = "Williams",
                          First = "Serena"),
              age = 36,
              record = c(wins = 801, losses = 136),
              grand.slam.wins = list(
                Australian.Open = c(2003, 
                                    2005, 
                                    2007, 
                                    2009,
                                    2010, 
                                    2015,
                                    2017),
                French.Open = c(2002,
                                2013,
                                2015),
                Wimbledon = c(2002,
                              2003,
                              2009,
                              2010,
                              2012,
                              2015,
                              2016),
                US.Open = c(1999,
                            2002,
                            2008,
                            2012,
                            2013,
                            2014)
                ))

print(list2)
```



##Question 2 - cbind, rbind
Create the following vectors and matrices.
```
## [1]  3  5  6 12
##
## [1]  TRUE FALSE  TRUE  TRUE
##
## [1] "A" "B" "C" "D" "E"
##
##      [,1] [,2] [,3] [,4] [,5]
## [1,]    2   14   26    5   17
## [2,]    5   17   29    8   20
## [3,]    8   20   32   11   23
## [4,]   11   23    2   14   26
##
##      [,1] [,2] [,3] [,4]
## [1,]    1   -1    1    0
## [2,]    0    0   -1    1
```
```{r}

num.vec <- c(3, 5, 6, 12)
log.vec <- c(TRUE, FALSE, TRUE, TRUE)
char.vec <- c("A", "B", "C", "D", "E")

mat1 <- matrix(data= c(2, 14, 26, 5, 17,
                       5, 17, 29, 8, 20,
                       8, 20, 32, 11, 23,
                       11, 23, 2, 14, 26),
               byrow=TRUE,
               ncol=5,
               nrow=4)

mat2 <- matrix(data = c(1, -1, 1, 0,
                        0, 0, -1, 1),
               byrow=TRUE,
               ncol=4,
               nrow=2)
cat("Created elements: \n\n")
cat("num.vec: \n", num.vec,"\n\nlog.vec: \n", log.vec,"\n\nchar.vec: \n", char.vec,"\nmat1: \n\n")
mat1
cat("\nmat2: \n\n")
mat2
```


###(a) cbind
Create the following three outputs using data objects created above and a single cbind() command (for each output).
```
##         v2
## [1,]  3  1
## [2,]  5  0
## [3,]  6  1
## [4,] 12  1
##      [,1] [,2]   
## [1,] "A"  "TRUE" 
## [2,] "B"  "FALSE"
## [3,] "C"  "TRUE" 
## [4,] "D"  "TRUE" 
## [5,] "E"  "TRUE"
##      [,1] [,2] [,3] [,4] [,5] [,6]
## [1,]    2   14   26    5   17    3
## [2,]    5   17   29    8   20    5
## [3,]    8   20   32   11   23    6
## [4,]   11   23    2   14   26   12
```


```{r, warning=FALSE}
# There has to be an easier way to do this.

cb1 <- cbind(num.vec, c(abs(mat2[,2]), abs(mat2[,3])), deparse.level=0)
cb1

# suppress errors because log.vec only had four values, but 
# repeating the first one gave use the "right" answer...
cb2 <- cbind(char.vec, log.vec, deparse.level = 0)
cb2


cb3 <- cbind(mat1, num.vec, deparse.level=0)
cb3
```


### (b) rbind
Create the following three outputs using data objects created above and a single rbind() command (for each output).
```
##      [,1] [,2] [,3] [,4]
## [1,]    2   14   26    5
## [2,]    5   17   29    8
## [3,]    8   20   32   11
## [4,]   11   23    2   14
## [5,]    1   -1    1    0
## [6,]    0    0   -1    1
##      [,1] [,2] [,3] [,4]
## [1,]    3    5    6   12
## [2,]    1   -1    1    0
## [3,]    0    0   -1    1
##    [,1]   [,2]    [,3]   [,4]   [,5]  
## v1 "3"    "5"     "6"    "12"   "3"   
## v2 "TRUE" "FALSE" "TRUE" "TRUE" "TRUE"
## v3 "A"    "B"     "C"    "D"    "E"
```

```{r, warning=FALSE}
cat("rb1:\n")
rb1 <- rbind(mat1[,1:4], mat2)
rb1

cat("\n\nrb2:\n")
rb2 <- rbind(num.vec, mat2, deparse.level = 0)
rb2

cat("\n\nrb3: \n")
rb3 <- rbind(as.character(num.vec), t(cb2))
rb3

```

##Question 3
###(a) Read in Iris data
Read in the csv Iris data directly from the UC-Irvine Machine Learning data repository into a dataframe (see iris.data link below).
•	Assign appropriate column names (see metadate on website).
•	Make sure columns have the correct data type. 
•	Use the summary() head() and/or View() functions to avoid dumping a large dataset into the console (and rmarkdown printout).
https://archive.ics.uci.edu/ml/machine-learning-databases/iris/iris.data
https://archive.ics.uci.edu/ml/machine-learning-databases/iris

```{r}
irisdf <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/iris/iris.data")
df_names <- readLines("https://archive.ics.uci.edu/ml/machine-learning-databases/iris/iris.names")

```
```{r}
colnames(irisdf) <- c("sepal_length_cm", "sepal_width_cm", "petal_length_cm", "petal_width_cm", "class")
irisdf$class <- as.factor(irisdf$class)
print(str(irisdf))

cat("\n Columns are now appropriate names and variable types.")

cat("\n", summary(irisdf), "\n\n\n")


```
```{r}
# view the dataset in another window
View(irisdf)
```

###(b) Present summaries (use summary function) of the following subsets
•	Iris data with class equal to Iris-versicolor.
•	Iris data with petal width greater than 1.5 cm OR sepal width greater than 3.0 cm.
–	Use |  compute a logical “OR” .  E.g. (x > 6  y > 2) gives all observations w/ x greater than 6 OR y greater than 2.

Do you see a relationship between the width measurements and class of Iris?

```{r}
cat("Summary of iris with clas == 'Iris-versicolor'\n")
summary(irisdf[irisdf$class == "Iris-versicolor",])
cat("\n\n")
cat("Summary of iris with petal_width_cm > 1.5cm OR sepal_width_cm > 3.0 cm.\n")
summary(irisdf[c(irisdf$petal_width_cm > 1.5 | irisdf$sepal_width_cm > 3.0), ])
cat("\n\n")
```

The classes viriginica and setosa exhibit wider petals or sepals than versicolor.

##Question 4 - Cars
###(a) Read in car data
Read in the car data data provided in the car.csv file. These data consists entirely of categorical data, so when reading the file set the stringsAsFactors parameter appropriately.
Use the summary() head() and/or View() functions to avoid dumping a large dataset into the console (and rmarkdown printout).

```{r}
car_df <- read.csv("week 3/car.csv", stringsAsFactors = TRUE)
head(car_df)
summary(car_df)
```

###(b) remove all observations (rows) with missing values
Note that is.na() can be used to identify NA values. Show dim and summary of data with appropriate observations removed.

```{r}

na_detector <- function(df_vec) {
  return(c(which(is.na(df_vec))))
}

rowNAs <- c(na_detector(car_df$price.buy),
            na_detector(car_df$price.maint),
            na_detector(car_df$doors),
            na_detector(car_df$seats),
            na_detector(car_df$trunk),
            na_detector(car_df$safety),
            na_detector(car_df$rating))

rowNAs

car_df2 <- car_df[-rowNAs,]

cat("dim before NA removal: ", dim(car_df), "\n\n dim after NA removal: ",dim(car_df2))
str(car_df2)

```

###(c) remove duplicate observations (rows)
Now remove all duplicate observations from NA-free car dataset. Show your results with the dim and summary functions.

```{r}
car_df3 <- car_df2[!duplicated(car_df2), ]
cat("dim of df with duplicates removed: ", dim(car_df3), "\n\n")
summary(car_df3)
```


