---
title: "Homework 1"
author: "Magnus Wood"
output: html_notebook
---
==

```{r}
plot(cars)
```

Question 1

(a)	Engine 1
Create the vector below.
##      Displacement (cc)     Net Power Out (kW)        Net Torque (Nm) 
##                 163.00                   3.60                  10.30 
##       Oil Capacity (l) Fuel Tank Capacity (l) 
##                   0.58                   3.10
Examine the attributes, dimension(s) and data type for this vector. Comment.

```{r "Q1s"}
# create a vector using key value pairs
my_vector <- c("Displacement (cc)" = 163, "New Power Out (kw)" = 3.60, "Net Torque (Nm)" = 10.30, "Oil Capacity (l)" = 0.58, "Fuel Tank Capacity (l)" = 3.10)

print(my_vector)


# Investigate the vector
vec_cls = class(my_vector)
vec_len = length(my_vector)
vec_attr = attributes(my_vector)

# print a formatted string with the information
writeLines(sprintf("class: % s,\nLength: % s,\nAttributes: % s", vec_cls, vec_len, vec_attr))
```

```{r "Q1b"}

     efficient_vector <- c("Displacement (cc)" = 163, "Net Power Out (kW)"=3.6, "Net Torque (Nm)"=10.3, "Oil Capacity (l)"=0.58, "Fuel Tank Capacity (l)"=3.1)

print(efficient_vector)

# Investigate the vector
vec_cls = class(my_vector)
vec_len = length(my_vector)
vec_attr = attributes(my_vector)

# print a formatted string with the information
writeLines(sprintf("class: % s,\nLength: % s,\nAttributes: % s", vec_cls, vec_len, vec_attr))

```



Question 2 – Family
Create a vector with names of your family members, making clear which name is for your father, wife, husband, son, cat, etc…

Question 3 - Matrices
(a)	Create the matrix below.
##        AA   BB   CC   DD   EE
## [1,] -3.3 -1.0  1.3  3.6  5.9
## [2,]  8.2 10.5 12.8 15.1 17.4
## [3,] 19.7 22.0 24.3 26.6 28.9
Use str to examine the properties of this matrix and comment.
(b)arithmetic with matrix
Add 3.3 to the matrix above, and divide by 2.3. Show your results and comment.
 (c)Create the matrix below.
Use prior results to complete this task as efficiently as possible.
##      [,1] [,2] [,3]
## [1,] -3.3 22.0 15.1
## [2,]  8.2  1.3 26.6
## [3,] 19.7 12.8  5.9
## [4,] -1.0 24.3 17.4
## [5,] 10.5  3.6 28.9
Use str to examine the properties of this matrix and comment.
 (d) Create the matrix below.
Use prior results to complete this task as efficiently as possible.
##      [,1] [,2] [,3]
## [1,]   -3   21   15
## [2,]    8    1   26
## [3,]   19   12    5
## [4,]   -1   24   17
## [5,]   10    3   28
Use str to examine the properties of this matrix and comment.



