---
title: "Homework 1"
author: "Magnus Wood"
output: html_notebook
---
==

### Question 1 (a)	Engine 1
Examine the attributes, dimension(s) and data type for this vector. Comment.

```{r "Q1a"}
car_vect <-  c("Displacement (cc)" = 163, "New Power Out (kw)" = 3.60, "Net Torque (Nm)" = 10.30, "Oil Capacity (l)" = 0.58, "Fuel Tank Capacity (l)" = 3.10)

names(car_vect) <- c("Displacement (cc)", "New Power Out (kw)", "Net Torque (Nm)", "Oil Capacity (l)", "Fuel Tank Capacity (l)")

print(car_vect)

# Investigate the vector
vec_cls = class(car_vect)
vec_len = length(car_vect)
vec_attr = attributes(car_vect)

# print a formatted string with the information
writeLines(sprintf("class: % s,\nLength: % s,\nAttributes: % s", vec_cls, vec_len, vec_attr))
```

```{r "Q1b"}

efficient_vector <- c("Displacement (cc)" = 163, "Net Power Out (kW)"=3.6, "Net Torque (Nm)"=10.3, "Oil Capacity (l)"=0.58, "Fuel Tank Capacity (l)"=3.1)

print(efficient_vector)

# Investigate the vector
vec_cls = class(efficient_vector)
vec_len = length(efficient_vector)
vec_attr = attributes(efficient_vector)

# print a formatted string with the information

cat("\nVector information: \nclass:", as.character(vec_cls), "\nLength: ", as.character(vec_len), "\nAttributes: ", as.character(vec_attr))

```



## Question 2 – Family

```{r}
#Create a vector with names of your family members, making clear which name is for your father, wife, husband, son, cat, etc…

my_family <-c("father" = "Nick", "mother" = "Mary", "brother" = "Davin", "sister" = "Bridget", "wife" = "Jocelyne", "son" = "Owen")

my_family
```

### Question 3 - Matrices
#(a)	Create the matrix below.
        AA   BB   CC   DD   EE
 [1,] -3.3 -1.0  1.3  3.6  5.9
 [2,]  8.2 10.5 12.8 15.1 17.4
 [3,] 19.7 22.0 24.3 26.6 28.9
 Use str to examine the properties of this matrix and comment.

```{r}

vec_data <- c(-3.3, 8.2, 19.7, -1.0, 10.5, 22.0, 1.3, 12.8, 24.3, 3.6, 15.1, 26.6, 5.9, 17.4, 28.9)
mat_data <- matrix(data = vec_data, nrow=3, ncol=5)
colnames(mat_data) <- c("AA", "BB", "CC", "DD", "EE")
mat_data

str(mat_data)

```
The data provided shows the shape of the matrix (3 rows and 5 columns), and lists two attributes available (dimensions, and the row and column names).The matrix is type numeric. 

### (b)arithmetic with matrix
Add 3.3 to the matrix above, and divide by 2.3. Show your results and comment.
```{r}

mat_data2 <- mat_data
mat_data2
mat_data2 <- mat_data2 + 3.3
mat_data2
mat_data2 <- mat_data2 / 2.3
mat_data2


```
Default operations appear to be element-wise. Works as expected. 

### (c)Create the matrix below.
Use prior results to complete this task as efficiently as possible.
Use str to examine the properties of this matrix and comment.

```{r}
mat_data3 <- matrix(data = vec_data, nrow = 5, ncol = 3)
mat_data3

str(mat_data3)

```

The str() function returns the shape of the matrix, plus the values in the matrix as a single vector. 


### (d) Create the matrix below.
Use prior results to complete this task as efficiently as possible.
Use str to examine the properties of this matrix and comment.

```{r}
mat_data4 <- matrix(as.integer(vec_data), nrow=5, ncol=3)
mat_data4[1, 2] = 21
str(mat_data4)
mat_data4
```
The matrix is of type numeric. Interesting, as the data I provided it was of type integer due to the use of as.integer(). Cool! I have mostly used the tidyverse dataframe, so using the base matrix was new to me. 


