---
title: "R Notebook Markdown fun"
subtitle: "A journey throught the remarkable world of cows"
author: "Magnus"
output: 
  html_document: html_notebook
  pdf_documnet: default

---

### Itemized list:
- Top level item 1
  - Bottom Level item 1
- Top level item 2
  - Bottom Level item 2

### Sentence with bold and italicized?
So how do I _write_ some *words* that are **bolded**. I guess this is how we __do that__?  
__NOW a whole *sentence*!__

### include and echo FALSE differences


```{r echo = FALSE, cache = TRUE}
# echo = False
vec = c("cow", "horse", "pig")
cat("echo = FALSE", vec)
```

```{r include = FALSE, cache = TRUE}
# include = False
vec = c("cow", "horse", "pig")
cat("include = FALSE\n", vec)
```

```{r}
# no False things here
vec = c("cow", "horse", "pig")
cat("nothing\n", vec)
```

include = FALSE doesn't display anything. Include is not displayed, and will not show results or the text. 

echo = FALSE doesn't display anything, either. 


### color font
A <span style="color:red">brilliant</span> <span style="color:green">world</span> of <span style="color:blue">color</span> awaits!

### Notes
* look into cache
* look into bookdown

