---
title: "AFS505 Module 1 Final Exam"
subtitle: "Part B"
date: "9/27/22"
author: "Magnus Wood"
output: html_document
---
```{r echo=FALSE}
library('rmarkdown')
library('bookdown')

```

## 4. Extra Credit (Optional)
### Create an R Markdown file with different tabs for each of the six parts of question 2.
### In a seventh tab 

## 2. You are provided a folder with three location (county) names, each of which has subfolders for one or two crops, which in turn has a data file.{.tabset}

### a. Iterate through the folders to read all the files and merge them into a single data frame. You can use a “loop” to iterate or for efficiency check out the list.files() function. 

```{r}
print("hello")
```

### b. Add four additional columns to the merged dataframe corresponding to the county name, crop name, latitude and longitude of the data. You must get this information from the directory structure you are looping through or the strings returned by the call to list.files().

```{r}
print("world")
```

### c. Rename the column irrig to irrigation_demand and precip to precipitation and export the dataframe as a csv file.
```{r}

```


### d. Summarize the annual irrigation demand by crop name and county name.
```{r}

```

### e. What is the average yield of Winter Wheat in Walla Walla at 46.03125N118.40625W for the year ranges (1981-1990), (1991-2000), and (2001-2019)?
###f. Which location has highest yield (average) for the time period (2001-2019) for grain corn?
```{r}

```

### g. Add the Github link which has your R script/ R markdown file and the csv file generated from 2 (c).

Please find all files related to this final exam in ["this Github."]
["Part A"](https://www.github.com/fieryWalrus1002/afs505_hw/)
["Part B : Question 2"](https://www.github.com/fieryWalrus1002/afs505_hw/)
["Data output of 2c"]

