---
title: "Combining data objects"
subtitle: "Data in R, Fall 2022"
output:
html_document: default
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Simple operations to combine data objects

Data wrangling packages like **dplyr** and **tidyr** can be used to reshape data, but it is often convenient to use base functions for simple data manipulations.

##cbind() and rbind() matrices

Let's start by created some demo data objects.
```{r matrix cbind vectors 1a}
mat1 <- matrix(1:12,4,3)
mat1

cat("\nvect1\n")
vect1 <- c(-1,0,NA,99)
vect1

cat("\nvect2\n")
vect2 <- c("bob","sue","ranjeev","mei")
vect2

cat("\nvect2\n")
vect3 <- c(T,F,F,T)
vect3
```

The **cbind()** or column-bind function can be used to *bind* a vector of data to a matrix.
```{r matrix cbind vectors 1b}

cat("\ncbind(mat1,vect1)\n")
cbind(mat1,vect1)

# we can avoid naming a column by extracting vector values
cat("\ncbind(mat1,vect1[1:4])\n")
cbind(mat1,vect1[1:4])

cat("\ncbind(vect1[1:4], mat1)\n")
cbind(vect1[1:4], mat1)

# if data types don't match, R will coerce
cat("\ncbind(mat1,vect2[1:4])\n")
cbind(mat1,vect2[1:4])

cat("\ncbind(mat1,vect3[1:4])\n")
cbind(mat1,vect3[1:4])
```

Note that coersion is always with the following precedence: character > numeric > logical. Order of data in cbind() doesn't matter.

If row numbers for data objects in cbind don't match, R will recycle or trim.
```{r matrix cbind vectors 2}

# R will recycle data if bound column isn't long enough
cat("\ncbind(mat1,vect1[1:3])\n")
cbind(mat1,vect1[1:3])    

cat("\ncbind(vect1[1:3], mat1)\n")
cbind(vect1[1:3], mat1) 

# R will trim if data is too long
cat("\ncbind(mat1,seq(1,12,by=2))\n")
cbind(mat1,seq(1,12,by=2))    # only first four values used in vector

cat("\ncbind(1:-5, mat1)\n")
cbind(1:-5, mat1)    # only first four values used in vector
```

Notice that the vector is recycled or trimmed to match the matrix, regardless of which comes first in the cbind() function.

We can also use cbind() with two (or more) matrices, or combinations of vectors and matrices.
```{r matrix cbind matrix}
cat("mat1\n")
mat1

cat("\nmat2\n")
# create 2nd matrix
mat2 <- matrix(-1:-8,4,2)
mat2

cat("\ncbind(mat1,mat2)\n")
cbind(mat1,mat2)

cat("\ncbind(mat1,vect1[1:4],mat2)\n")
cbind(mat1,vect1[1:4],mat2)

#what happens if number of matrix rows don't match?
#cat("\ncbind(mat1,t(mat2)\n")
#cbind(mat1,t(mat2))
```

The **rbind()** or row-bind function similarly binds vectors and/or matrices row-wise.

Let's start by creating three slightly different data objects.
```{r matrix and vector create}
mat1 <- matrix(1:12,3,4, byrow=F)
cat("\nmat1\n")
mat1

mat2 <- matrix(-1:-8,2,4)
cat("\n\nmat2\n")
mat2

vect1 <- c(-1,0,NA,99)
cat("\n\nvect1\n")
vect1
```

Now let's use rbind.
```{r matrix rbind and cbind 1b}

cat("\n> rbind(mat1,mat2)\n")
rbind(mat1,mat2)

cat("\n> rbind(mat1,vect1)\n")
rbind(mat1,vect1)

cat("\n> rbind(mat1,vect1[1:4])\n")
rbind(mat1,vect1[1:4])

cat("\n> rbind(mat1,vect1[1:3])\n")
rbind(mat1,vect1[1:3])

cat("\n> rbind(mat1,1:5)\n")
rbind(mat1, 1:5)

#cat("\n> rbind(mat1,t(mat2)\n")
#rbind(mat1,t(mat2))
```

What happens if we bind two vectors using cbind() or rbind()?
  ```{r bind vectors using cbind rbind}
cat("cbind(1:5,-5:-9)\n")
cbind(1:5,-5:-9)

cat("\nrbind(1:5,-5:-9)\n")
rbind(1:5,-5:-9)

```

##cbind() and rbind() dataframes

The **cbind()** function operates slightly differently for dataframes.

Let's recreate the student dataframe discussed previously.
```{r student dataframe cbind 1}
name <- c("Leila", "Vijay", "Shira", "Behrooz")
gender <- factor(c("F","M","F","M"))
age <- c(18,23,48,19)
height <- c(60,70,65,72)
grade <- factor(c("A", "D", "B","C"), levels=c("F","D","C","B","A"), ordered=T)

# Now put all of these factors and vectors of same length into a data frame.
students <- data.frame(name, gender, age, height, grade)
students
```

Now let's add a variable for major.
```{r student dataframe cbind2}

cbind(students,c("soil sci", "hort", "econ", "entomology"))

cbind(students,major = c("soil sci", "hort", "econ", "entomology"))

```

Now let's examine the difference between the cbind and data.frame functions.
```{r student dataframe cbind3}

major <- c("soil sci", "hort", "econ", "entomology")

cbind(students,major)

data.frame(students,major)

# cbind applied to dataframes is really just a wrapper function for data.frame. 
# You are better off using data.frame due to greater functionality.
```

It is good practice to use data.frame instead of cbind to join columns of data in data frames.

We can also use rbind() with data frames to add rows (observations).
```{r student dataframe rbind1}
students <- data.frame(name, gender, age, height, grade)
students

students2 <- data.frame(name="Bob", gender="M", age=22, height=69, grade=factor(c("F"), levels=c("F","D","C","B","A"), ordered=T))
students2

rbind(students, students2)
```

Be careful about how R will coerce data if the data types in a given column don't match.
```{r student dataframe rbind2}

students2 <- data.frame(name="Bob", gender="M", age=22, height=69,grade="F")
students2

rbind(students, students2)
str(rbind(students, students2))
```

In this case, R converted the character "F" in *student2* to the ordinal value "F" in the first dataframe.

What happens if column names don't match?
```{r student dataframe rbind3}

# notice Age is capitalized for students2 vs lower case for students.
students2 <- data.frame(name="Bob", gender="M", age=22, height=69, grade="F")
students2

rbind(students, students2)
```

# Advanced addressing - dataframes

read wine
```{r read wine}
wine.dat <- read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data',header=F)
colnames(wine.dat) <- c("class","alcohol","malic.acid","ash","alkalinity.ash","mg","tot.phenol","flavanoids","nonflav.phenol","proanthocyanins","col.intensity","hue","OD280.OD315","proline")
wine.dat$class<-as.factor(wine.dat$class)
summary(wine.dat)
str(wine.dat)
```

Now get subset with proline < 1500.

```{r}
wine.dat2 <- wine.dat[wine.dat$proline < 1500, ]
summary(wine.dat2)
```

How do we select class == 1?
```{r eval=FALSE}
wine.dat3 <- wine.dat[wine.dat$class==1,]
summary(wine.dat3$class)

cat("\n")
wine.dat3 <- wine.dat[wine.dat$class=="1",]
summary(wine.dat3$class)

cat("\n")
wine.dat3 <- wine.dat[wine.dat$class==factor("1", levels=c("1","2","3")),]
summary(wine.dat3$class)

cat("\n")
wine.dat3 <- wine.dat[wine.dat$class==factor("1"),]
summary(wine.dat3$class)

```

## unique or duplicated vector values

The unique() and duplicated() functions are often useful.
```{r vectors unique duplicated}
vect1 <- c(0,1,1,2,1,3,4)
vect1

#which positions are duplicated?
cat("\nduplicated positions\n")
duplicated(vect1)

#remove duplicated positions
cat("\nremove duplicated positions\n")
vect1[!duplicated(vect1)]

# can also be accomplished with unique()
cat("\nunique values\n")
unique(vect1)
```

These functions can be extended to dataframes.

## unique or duplicated rows

Create students dataframe with duplicate rows.
```{r}
name <- c("Leila", "Vijay", "Shira", "Behrooz")
gender <- factor(c("F","M","F","M"))
age <- c(18,23,48,19)
height <- c(60,70,65,72)
grade <- factor(c("A", "D", "B","C"), levels=c("F","D","C","B","A"), ordered=T)
students <- data.frame(name, gender, age, height, grade)
students2 <- rbind(students,students[2,])
students2
```

Now apply duplicates, unique. Note that these are applied by default to entire rows.
```{r dataframe dup unique}
# duplicated is applied row-wise to dataframes

cat("\nduplicated rows in students2\n")
duplicated(students2)

# remove duplicated rows in students2
students2[!duplicated(students2),]

# select unique rows in students2
unique(students2)
```

Where duplicated() comes in useful is when you are only concerned about duplicates in certain columns of a dataframe, and want to remove rows where those columns are duplicated.
```{r dataframe dup unique 2}
#duplicated is applied row-wise to dataframes
students3 <- students2
students3$grade[2] <- "C"
students3

cat("\nduplicated rows in students3[,1:4]\n")
duplicated(students3[,1:4])

# remove duplicated rows in students3
students3[!duplicated(students3[,1:4]),]

#unique rows in students3
unique(students3)

#unique rows in unique(students3[,1:4])
unique(students3[,1:4])
```

Duplicated removes the row where the first four columns are duplicated, whereas unique can only be applied to an entire row.

##merge()

We can use the merge() function to join two data frames connected by one or more index columns. This will be addressed in greater depth in data wrangling.
```{r student dataframe cbind}
students3

name2 <- c("Leila","Vijay","Bob","Behrooz")
major <- c("soils","hort","plant path","econ")
students.m <- data.frame(name=name2,major)
students.m

#default merge, assume index columns have same names (name column)
merge(students3,students.m)

#change merge parameters. all.x keep all x dataframe rows? all.y keep all y dataframe rows? Sort result?
merge(students3,students.m, all.x=T, all.y=F, sort=F)
```

#Addressing data in functions

There are multiple ways to use data in a function.

First let's create a dataframe to use as an example.
```{r create dataframe}
mydat <- data.frame(x.var=c(1,5,7,9, 10),y.var=c(2,10,20,90,85))
mydat
```

##explicitly address all variables

plot variables addressed.
```{r data explicit}
lm(mydat$y.var ~ mydat$x.var)
plot(mydat$x.var, mydat$y.var)
```

plot variables addressed with column indexing
```{r}
lm(mydat[,2] ~ mydat[,1])
plot(mydat[,1], mydat[,2])
```

Plot with data =
  ```{r data equals}
#lm(y.var ~x.var)
#lm(y.var~x.var, data=mydat)
#plot(x.var, y.var, data=mydat)
```

Doesn't work for all functions.

Use with() function.
```{r with}
with(mydat, lm(y.var ~x.var))
with(mydat, plot(x.var, y.var))
```

Use attach data (warning, be careful, detach after use).
```{r attach}
attach(mydat)
lm(y.var ~x.var)
plot(x.var, y.var)
detach(mydat)
```